Testing Statistical Graphics

### 2022-07-05 index.qmd
couple of comments

There are heuristics, rules-of-thumb, and various guidelines \cite{asme-standards-graphics,allenVisualizingScientificData2016,haemerDoubleScalesAre1948,fewInformationDashboardDesign2006,kosslynGraphDesignEye2006} for creating useful and visually attractive charts, but as scientists, we believe empirical evidence is essential for making correct decisions. 
The best way to establish the efficacy of various design decisions, then, is to test our charts by having them evaluated by people under controlled conditions and gathering appropriate data\cite{clevelandShapeParameterTwoVariable1988,clevelandGraphicalPerceptionGraphical1985}. 
It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart\cite{eellsRelativeMeritsCircles1926,croxtonBarChartsCircle1927,vonhuhnFurtherStudiesGraphic1927,croxtonGraphicComparisonsBars1932}, but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally impossible to complete.
In this paper, we attempt to distill lessons learned from several different sets of experimental graphics experiments\cite{clusters,sineillusion,emily-diss,power,vanderplasFramedReproducingRevisiting2019}, discussing the use of several different testing methods for statistical graphics\cite{vanderplasTestingStatisticalCharts2020}, the process of designing a graphical experiment, and analysis of the resulting empirical data. 
It is our hope that by writing this paper, we will prevent other researchers from repeating hard lessons in what to do (and what not to do), lowering the barriers that exist for conducting research involving empirical tests of statistical graphics.

@@

Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power}. 
As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot instead of any of the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a p-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}.
In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models. This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hiearchy of visual attributes\cite{clusters}.
Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart. 
An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log.

@@

Of course, as we have confined ourselves to the discussion of online, asynchronous testing methods, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information. 
While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.
In addition, we have also provided participants with supportive tools, such as "scratch pads" and limited function calculators built into the Shiny application that allow us to gain additional insight into the estimation and visual comparison process when more complex calculations are required to answer the numerical estimation question (@fig-estimation-calc). 
These supportive tools have provided a wealth of information that, while not easy to analyze and generally messy, provides real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.

@@

It is also possible to design a task that requires participants to *engage directly* with an interactive visualization. 
This can be useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner. 
In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions you can ask in testing graphics is in your ability to write code to interact with the visualization library. 
This type of testing method can be extremely natural for participants, but it also is hard to generalize when discussing testing methods because of the potential range of applications where it might be employed.

>>>>

There are heuristics, rules-of-thumb, and various guidelines \cite{asme-standards-graphics,allenVisualizingScientificData2016,haemerDoubleScalesAre1948,fewInformationDashboardDesign2006,kosslynGraphDesignEye2006} for creating useful and visually attractive charts, but as scientists, we believe empirical evidence is essential for making correct decisions. 
The best way to establish the efficacy of various design decisions, then, is to test our charts by having them evaluated by people under controlled conditions and gathering appropriate data\cite{clevelandShapeParameterTwoVariable1988,clevelandGraphicalPerceptionGraphical1985}. 
It can be extremely challenging to design studies which strike the right combination of internal and external validity. 
Simply asking people to read quantities off of a graph may not generalize beyond the questions asked or the data used in the chart\cite{eellsRelativeMeritsCircles1926,croxtonBarChartsCircle1927,vonhuhnFurtherStudiesGraphic1927,croxtonGraphicComparisonsBars1932}, but designing a study that is sufficiently robust to those issues requires manipulation of so many factors as to be functionally close to impossible to complete.
In this paper, we attempt to distill lessons learned from several different sets of experimental graphics experiments\cite{clusters,sineillusion,emily-diss,power,vanderplasFramedReproducingRevisiting2019}, discussing the use of several different testing methods for statistical graphics\cite{vanderplasTestingStatisticalCharts2020}, the process of designing a graphical experiment, and analysis of the resulting empirical data. 
It is our hope that by writing this paper, we will prevent other researchers from repeating hard lessons in what to do (and what not to do), lowering the barriers that exist for conducting research involving empirical tests of statistical graphics.

@@

Statistical lineups\cite{bujaStatisticalInferenceExploratory2009,wickhamGraphicalInferenceInfovis2010} are a more generally applicable testing tool for perceptual questions such as "which chart displays this data more clearly"\cite{power} while simultaneously assessing the significance of the graphical finding in the chart. 
As in the criminal procedure, lineups embed a target plot with signal into an array of 19 other innocent "null" plots that show only noise; if viewers consistently pick the target plot at a higher rate than the 19 null plots, then the target plot is said to be visually significant\cite{majumderValidationVisualStatistical2013,vanderplasStatisticalSignificanceCalculations2021} and a "see" value (the visual analogue of a $p$-value for a statistical test) can be calculated\cite{niladriroychowdhurySeeValueApp2020}. 
<!-- Beyond the quantitative assessment of the signal with $p$- and see-values, the human involvement also allows us to collect information on the qualitative nature of *what* a viewer sees.--> 
In another variation of the statistical lineup procedure, two models are compared, with target plots from each model embedded in the array of 20 total plots; the remaining null plots are constructed from a mixture model that blends the two competing models. This variation allows the experimenter to assess graphical design choices to determine whether they effectively emphasize structural differences in the data and assign a hiearchy of visual attributes\cite{clusters}.
Statistical lineups do not require the experimenter to provide any contextual information for the data: all of the necessary information is embedded in the choice of a null statistical model, which is convenient for testing purposes, but does not require engagement at a level that would allow for testing the viewer's understanding of the information shown in the chart. 
An example of a statistical lineup for testing the perception of exponential growth is shown in @fig-lineup-log.

@@

Of course, as we have confined ourselves to the discussion of online, asynchronous testing methods, every user interaction with the testing materials (typically hosted on a web page) can also be recorded along with time stamps, mouse positions, browser size and screen resolution, and other information. While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question. What we typically see, is that the first round of test questions takes the longest for participants to complete. Additional experiments do not affect acuracy (i.e. there is no immediate learning effect) until after `too many' tests cognitive fatigue proves to be detrimental to accuracy\cite{chowdhury2018}. This sweet point between user replicates and fatigue depends on the cognitive burden in each test and should factor into setting the experiment.
<!--While we have not used this type of information heavily in our experimental analyses thus far, we do collect some time stamp data in order to assess how long participants spend on each question; it could almost certainly be more informative in the right setting.-->
In addition, we have also provided participants with supportive tools, such as "scratch pads" and limited function calculators built into the Shiny application that allow us to gain additional insight into the estimation and visual comparison process when more complex calculations are required to answer the numerical estimation question (@fig-estimation-calc). 
These supportive tools have provided a wealth of information that, while not easy to analyze and generally messy, provides real insight into how participants were looking at the data, what strategies they tried and discarded for reading the chart, and what visual estimation methods were used.

@@

It is also possible to design a task that requires participants to *engage directly* with an interactive visualization. 
This can be useful in a directed task, where users are asked to interact with the chart in a specific way and the result is recorded, but it is also possible to use interactive visualizations in an open-ended task, recording how users engage with the graphic in an exploratory (as opposed to goal-directed) manner. 
In one recent experiment, we asked participants to forecast an exponential trend, with data presented on either a linear or log scale. 
Using JavaScript code modified from New York Times interactive graphics "You Draw It" features\cite{katzYouDrawIt2017}, we had users draw trend lines with their computer mouse and make forecasts directly on interactive charts, with the data and user-drawn predictions recorded to our database.
With interactive graphics rendered using JavaScript (or other web libraries), the only limit to the types of questions one can ask in testing graphics is one's ability to write code to interact with the visualization library. 
This type of testing method can be extremely natural for participants, but it also is hard to generalize when discussing testing methods because of the potential range of applications where it might be employed.
